A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)

### 2018-04-13 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
updates

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed through SVD to principal component spaces where usually, traditional principal component analysis is performed before interpolation is used to implement further data reduction in the principal component spaces. This reduced data in the principal component space is then transformed back into the permeability space to obtain a reduced permeability space.

The principal component space is more challenging to visualize because it usually consists of more than $3$ dimensions. $3-D$ ({\it i.e.}, $x, y$ and $z$ in the cartesian plane) being the natural dimension which humans exist, is easy to visualize. Many individuals are also able to understand the concept of 4D with time as the fourth dimension, albeit only few scientists and engineers are able to visualize and work in the abstract dimension where the number of dimensions exceed four. An aid for visualizing the PC space is to write out the $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices (or imagine the matrices written out) and consider the singular vectors ({\it i.e.},columns in $\mathbf{U}$ and rows in $\mathbf{U}$) as separate principal component spaces.

\subsubsection{Model Order Reduction/Principal Component Space Reduction}\label{subsubsection:svdcase_preprocess_algorithm}
 For the SVD model order reduction of the permeability field, all the work is done in the principal component spaces. This can be completed in two steps. The first step is optional and it is the well known principal component analysis (PCA) \cite{Hotelling_1933}. Here, the singular values are examined and based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$. In the case with $z = r$, it can be assumed that step 1 has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .

The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$), to reduce the number of data in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

\subsection{Model Description}\label{subsection:model}
The BaseCase  ({\it i.e.} fine resolution) model for the permeabiliy field builds on the $2 \times 2$ block, $4$ quadrant model used initially by \citet{Cardwell_1945} and later by \citet{Yeo2001} and \citet{dawe_2008}. These authors designed the permeability field for the quadrants in a checkerboard style which means that diagonally opposite blocks retained the same permeability values with one set of diagonally opposite blocks having a higher permeability value compared to the other set. Hence, one set of diagonally opposite blocks are referred to as high permeability blocks while the other set is referred to as low permeability blocks. The BaseCase permeability model used in this manuscript retains the checkerboard desgined used by the recently named authors. Albeit, the permeability data for each block in the set of diagonally opposite blocks have similar value ({\it i.e.} not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. And Fig.~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible respectively. Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right respectively.

>>>>

\subsubsection{Understanding the principal component spaces}\label{subsubsection:visualization_pcspaces}
Essentially, the existing permeability space is transformed through SVD to principal component spaces (PCS) where traditional principal component analysis is performed before interpolation is used to implement further data reduction in the principal component spaces. This reduced data in the PCS is then transformed back into the permeability space to obtain a reduced permeability space.

The PCS is more challenging to visualize because it usually consists of more than 3-D. An aid for visualizing the PCS is to write out the $\mathbf{U}$, $\mathbf{\Sigma}$ and $\mathbf{V^{\intercal}}$ matrices and consider the singular vectors (\ie columns in $\mathbf{U}$ and rows in $\mathbf{U}$) as separate PCS.

\subsubsection{Model Order Reduction/Principal Component Space Reduction}\label{subsubsection:svdcase_preprocess_algorithm}
For the SVD model order reduction of the permeability field, all the work is done in the principal component spaces. This can be completed in two steps: the first step is the principal component analysis (PCA) \cite{Hotelling_1933}, where the singular values are examined and, based on a pre-defined criteria, $z$ singular values are selected, starting from $\sigma_{1}$ and in a decreasing order, where $z \leq r$.

In the case with $z = r$, it can be assumed that step 1 has been ignored. However, when $z < r$, the first $z$ columns of $\mathbf{U}$ and the first $z$ rows of $\mathbf{V^{\intercal}}$ are extracted from $\mathbf{U}$ and $\mathbf{V^{\intercal}}$ to form new matrices $\mathbf{U_{new}}$ and $\mathbf{V^{\intercal}_{new}}$ respectively. The selected $z$ singular values also form a new diagonal matrix, $\mathbf{\Sigma_{new}}$ .

The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) to reduce the data in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

\subsection{Model Description}\label{subsection:model}
The BaseCase  (\ie fine mesh resolution) model for the permeabiliy field builds on the $2 \times 2$ block, $4$ quadrant model used initially by \citet{Cardwell_1945} and later by \citet{Yeo2001} and \citet{dawe_2008}. These authors designed the permeability field for the quadrants in a checkerboard style which means that diagonally opposite blocks retained the same permeability values with one set of diagonally opposite blocks having a higher permeability value compared to the other set. Hence, one set of diagonally opposite blocks are referred to as high permeability blocks while the other set is referred to as low permeability blocks. The BaseCase permeability model used in this manuscript retains the checkerboard desgined used by the recently named authors. Albeit, the permeability data for each block in the set of diagonally opposite blocks have similar value ({\it i.e.} not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. And Fig.~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible respectively. Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right respectively.
